---
title: "Bitcoin Price Forecasting"
author: "Keerthikant"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#####  The data we have considered is from Apr 23 2013 to May 9, 2019. I have extracted the data from the website (https://coinmarketcap.com/).

##### Forecast is a generic function for forecasting from time series or time series models. The function invokes particular methods which depend on the class of the first argument.

##### Naniar package is used to plot heatmap to check missing values in the data

##### Lubridate package is used for to work with dates and times

```{r echo=TRUE, message=FALSE, warning=FALSE}
setwd('C:/Users/keert/Documents/Keerthi/Personal/ML/Time series/bitcoin/Code')
library(forecast)
library(ggplot2)
library(lubridate)
library(dplyr)
library(naniar)
library(tseries)
```

##### Let's load the input file which is the training file provided in the bitcoin dataset.

```{r echo=TRUE}
train_data=read.csv('C:/Users/keert/Documents/Keerthi/Personal/ML/Time series/bitcoin/Data/Bitcoin_Train_Data.csv',header=TRUE)
test_data=read.csv('C:/Users/keert/Documents/Keerthi/Personal/ML/Time series/bitcoin/Data/Bitcoin_Test_Data.csv',header=TRUE)

```

##### Initial checks to convert date field so that we can work on time series models.

```{r echo=TRUE}
str(train_data)

train_data$Date <- mdy(train_data$Date)
test_data$Date <- mdy(test_data$Date)

train_data$Volume = gsub('-','0',train_data$Volume)
train_data$Volume = as.numeric(train_data$Volume)

```

##### Initial summaries to see any outliers in any of the columns

```{r echo=TRUE}
head(train_data)
summary(train_data)
```

##### Plotting heatmap to check for missing data

```{r echo=TRUE}
#Heat map to check any missing data
vis_miss(train_data)
```

* The data is clean. There are no missing element. There is nothing unusual about the data.

##### We plot the initial bitcoin close price for each day from Apr 2013 to May 2019.

```{r echo= TRUE}
plot(train_data$Close~train_data$Date,xlab='Date',ylab ='Daily Bitcoin Close Price in USD',
     main='Bitcoin price plot', type='l',lwd=2,col='blue')
```

### We proceed with next steps to make the data stationary to apply time series models -
* astsa package is used for applied stastical time series models 
* Creating timeseries data using ts()

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(astsa)

bit_train<-ts(train_data$Close)
bit_test<-ts(test_data$Close)

```

```{r echo=TRUE}

par(mfrow=c(2,2))

plot(bit_train, main='Daily Close Prices of Bitcoin in USD', ylab='', col='blue', lwd=2)
plot(log(bit_train), main='Log-transorm of Close Prices', ylab='', col='red', lwd=2)
plot(diff(log(bit_train)), main='Differenced Log-transorm of Close Prices', ylab='', col='brown', lwd=2)
plot(diff(diff(log(bit_train)),12), main='Log-transorm without trend and seasonaliy', ylab='', col='green', lwd=2)

```

* We notice that our data is non-stationary, we apply log-return method to make it stationary.
* Log return is achieved by diff(log())
* We plot 4 graphs in efforts to see if log-return methods makes the data stationary before we apply AR or MA or ARMA or ARIMA models.

##### Performing Ljung-Box test for correlation

```{r echo=TRUE}
Box.test(diff(log(bit_train)), lag = log(length(diff(log(bit_train)))),type ='Ljung-Box')
```

* We can notice that p < 0.05, there exists some correlation with the close price of the bitcoin

##### We perform dickey-fuller test to determine stationary in the data

```{r echo=TRUE}
adf.test(diff(log(bit_train)), alternative="stationary", k=0)
```

* We can conclude that there is no stationarity from above p-value

##### Auto Correlation Function (ACF) and Partial Auto Correlation Function (PACF) of the differenced data is observed to get the p,d,q values.

```{r echo=TRUE}

par(mfrow=c(1,2))
acf(diff(log(bit_train)), main='ACF of log differenced data', 100)
pacf(diff(log(bit_train)), main='PACF of log differenced data', 100)

```

##### We fit arima(4,1,3) model with seasonal component included in it, below plot shows the predictions

```{r echo=TRUE}

fit <- arima(log(bit_train), c(4, 1, 3),seasonal = list(order = c(1, 1, 1), period = 3))
pred <- predict(fit, n.ahead = 10*3)
ts.plot(bit_train,2.718^pred$pred, lty = c(1,5),col=c(3,2),lwd=c(2,3))
grid(nx = NULL, ny = NULL, col = "darkgray", lty = 2,lwd = par("lwd"), equilogs = TRUE)

```

* We have also used auto.arima() to check the p,d,q which complements the ACF and PACF graphs

### We now proceed with Exponential Smoothing Methods
##### 1. Simple Exponential Smoothing  
##### 2. Holt's Linear Trend Method  
##### 3. Damped Holt's Method  

* We have reduced the window size for the time series to start from observation number= 2000 for graph plotting purposes.

##### We notice that AIC is 3122 and MAPE=2.009 for Simple Exponential Smoothing

```{r echo=TRUE}
close_ses <- ses(window(bit_train,start=2000),h=5)
summary(close_ses)

autoplot(close_ses) +
  autolayer(fitted(close_ses), series="Fitted") +
  ylab("Bitcoin Close Price") + xlab("ID number")
```

##### We notice that AIC is 3123 and MAPE=2.024 for Holt's method

```{r echo=TRUE}
close_holt<-holt(window(bit_train,start=2000),h=5)

summary(close_holt)

autoplot(close_holt) +
  autolayer(fitted(close_holt), series="Fitted") +
  ylab("Bitcoin Close Price") + xlab("ID number")
```

##### We notice that AIC is 3121 and MAPE=1.9867 for Damped Holt's method. This has the lowest MAPE as compared to above methods.

```{r echo= TRUE}
close_holt_damp<-holt(window(bit_train,start=2000),h=5,damped = TRUE,phi=0.9)

summary(close_holt_damp)

autoplot(close_holt_damp) +
  autolayer(fitted(close_holt_damp), series="Fitted") +
  ylab("Bitcoin Close Price") + xlab("ID number")

```

##### Plotting in Holt's linear method forecasts and Damped Holt's method  forecasts in single graph.

```{r echo=TRUE}
autoplot(window(bit_train,start=2180)) +
  autolayer(close_holt, series="Holt's method", PI=FALSE) +
  autolayer(close_holt_damp, series="Damped Holt's method", PI=FALSE) +
  ggtitle("Forecasts from Holt's method") + xlab("ID Number") +
  ylab("Bitcoin Price in USD") +
  guides(colour=guide_legend(title="Forecast"))

```

### We now proceed to next stage: determing the model parameters using ets() E= Error, T=Trend, S=Seasonal

```{r echo=TRUE}
bit_train_2000 <- window(bit_train, start=2000)
model_param <- ets(bit_train_2000,model="ZZZ",damped = TRUE)
summary(model_param)
```

* We see alpha=0.895, beta=0.0825, phi=0.9073 
* The small values of beta mean that the slope and seasonal components change very little over time. The
narrow prediction intervals indicate that the series is relatively easy to forecast
due to the strong trend and seasonality.  

```{r echo=TRUE}
autoplot(model_param)
model_param %>% forecast(h=8) %>%
  autoplot() +
  ylab("Bitcoin Price in USD")
```

